Testing Gravity Using Type Ia Supernovae Discovered by Next Generation Wide Field Imaging Surveys

### 2019-01-19 doc/article.tex
updating

\documentclass{aastex62}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{graphicx}				% Use pdf, png, jpg, or epsÂ§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb,amsmath}
\usepackage{natbib,verbatim}
\usepackage{grffile}
\usepackage{textcase}
\usepackage{hyperref}
%SetFonts

@@

\clearpage
\newpage
\setcounter{page}{0}
\pagenumbering{arabic}
\setcounter{page}{1}
\title{Testing Gravity Using Type Ia Supernovae Discovered by Next Generation Wide Field Imaging Surveys}
\author[0000-0001-6315-8743]{A.~G.~Kim}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{P.~Antilogus}
\affiliation{Laboratoire de Physique Nucl\'eaire et de Hautes Energies, Sorbonne Universit\'e, CNRS-IN2P3, 4 Place Jussieu, 75005 Paris, France}    
\author{S.~BenZvi}
\affiliation{Department of Physics and Astronomy, University of Rochester, Rochester, NY 14627, USA}
\author{S.~Gontcho A Gontcho}
\affiliation{Department of Physics and Astronomy, University of Rochester, Rochester, NY 14627, USA}
\author{R.~Graziani}
\affiliation{Universit\'e Clermont Auvergne, CNRS/IN2P3, Laboratoire de
Physique de Clermont, F-63000 Clermont-Ferrand, France}    
\author{C.~Harper}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{C.~Howlett}
\affiliation{International Centre for Radio Astronomy Research, The University of Western Australia, Crawley, WA 6009, Australia}
\author{D.~Huterer}
\affiliation{Department of Physics, University of Michigan, 450 Church Street, Ann
Arbor, MI 48109, USA }
\author{C.~Ju}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{P.-F.~Leget}
\affiliation{Laboratoire de Physique Nucl\'eaire et de Hautes Energies, Sorbonne Universit\'e, CNRS-IN2P3, 4 Place Jussieu, 75005 Paris, France}    
\author{E.~V.~Linder}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{P.~McDonald}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{J.~Nordin}
\affiliation{Institut fur Physik, Humboldt-Universitat zu Berlin, Newtonstr. 15, 12489 Berlin, Germany}
\author{S.~Perlmutter}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\affiliation{Department of Physics, University of California Berkeley, 366 LeConte Hall MC 7300, Berkeley, CA, 94720-7300}
\author{N.~Regnault}
\affiliation{Laboratoire de Physique Nucl\'eaire et de Hautes Energies, Sorbonne Universit\'e, CNRS-IN2P3, 4 Place Jussieu, 75005 Paris, France}    
\author{M.~Rigault}
\affiliation{Universit\'e de Lyon, F-69622, Lyon, France; Universit\'e de Lyon
1, Villeurbanne; CNRS/IN2P3, Institut de Physique Nucléaire de
Lyon, France}    
\author{others}

@@

\begin{abstract}
ZTF today and LSST in the upcoming decade will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands.  The increase in the number density of SNe~Ia, in parallel with improvements in the standardization of
their absolute magnitudes, now make them competitive probes of the growth of structure.  The peculiar velocity power spectrum
is sensitive to $f\sigma_8$,
the product of the linear growth and amplitude of density perturbations.  Cross-correlation with synergistic galaxy surveys further constrains $f \sigma_8$ and the galaxy bias. 
Thus, in the next decade the peculiar velocities of
SNe~Ia will provide the growth of structure  in the local $z<0.3$ Universe as a powerful test of General Relativity and other models of gravity.
\end{abstract}

\section{Connection between Type Ia Supernovae Correlations and Gravity}

@@

The growth of structure depends on the expansion history of the Universe, the nature and density of its contents, and gravity. 
It is therefore a powerful probe of cosmology and dark energy.  Growth of structure can be measured from the light-emittting structures
in the Universe and from the peculiar velocities of test masses therein.
Peculiar velocities are the motions, on top of the cosmological expansion, caused by the gravitational attraction
and repulsion of density inhomogeneities in the Universe.  The peculiar velocity of an object with a known absolute magnitude
can be determined from its observer-frame magnitude and redshift. For a given background cosmology, the observed magnitude provides
an estimate of the cosmological redshift, the peculiar velocity is then the difference between the cosmological and CMB-frame redshifts.

Baryonic structures and peculiar velocities  provide a measurement of the combination $fD$.  $D$ is the ``linear growth factor'' that
gives the overall amplitude of  overdensities, and the ``linear growth rate''
$$f \equiv \frac{d\ln{D}}{d\ln{a}}$$ is how that amplitude changes with redshift.  General Relativity predicts
$f \approx \Omega_M^\gamma$ (with $D$ determined accordingly) with $\gamma=0.55$, whereas other gravity models can be similarly described
with different values of $\gamma$
\citep{2007APh....28..481L}.  
The growth of structure, through the measurement of $fD$, provides a test of General Relativity and breaks degeneracies
between gravitational and dark energy models that explain the accelerating expansion of the Universe.
Non-GR models may  also predict a change in the redshift- and scale-dependence of the growth, such observations provide additional leverage in probing gravity.
The parameter $\sigma_8$, the  standard deviation of overdensities in 8$h^{-1}$Mpc spheres, is 
commonly used in place of $D$ to normalize the
overall amplitude of  overdensities, so the standard parameterization used by the community is $f\sigma_8$.

Baryonic structures are sensitive to $f\sigma_8$ through redshift space distortions (RSD), 
which on large scales gives the combination $(b + f \mu^2)\sigma_8$ where $b$ is the bias between the tracer and dark matter
and $\mu$ gives the angular separation of Fourier modes to the line-of-sight.  Correlations between peculiar velocities are sensitive to  $f\sigma_8 (H_0 d_L)^{-1}$.
Although the velocities may be measured off of biased tracers, the tracers' dynamics are driven by all mass (including dark matter) and so are bias-free.
As
mentioned earlier, peculiar velocities are measured relative to the background cosmological expansion that
leads to the dependence on $H_0 d_L$. (Note that the combination $H_0 d_L$ is independent of $H_0$.)
Baryonic structures and peculiar velocities within the same volume are induced by the same overdensities  making their cross-correlation 
insensitive to sample variance \citep{2007PhRvL..99h1301G}: galaxy and peculiar velocity surveys provide synergistic constraints more powerful
than their naive sum.

The probative power of a specific peculiar-velocity tracer primarily depends on its number density and the precision to which its absolute magnitudes are known.
The current generation of peculiar velocity studies uses $10^3-10^5$ galaxies with Fundamental Plane and Tully-Fisher distances \citep{2008AJ....135.1738M, 2014MNRAS.445.2677S,
2016AJ....152...50T}.  These galaxies have absolute magnitude uncertainties of $\sim 0.4$~mag. 
Next generation surveys WALLABY \citep{2008ExA....22..151J} and TAIPAN \citep{2017PASA...34...47D} are designed to increase these sample sizes by an order of magnitude
over 75\% of the sky to a maximum depth of $z=0.1$.

@@

Two advances in the upcoming decade will make SNe~Ia  important probes of $f\sigma_8$.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields $\sigma_M \gtrsim 0.12$~mag absolute magnitude
dispersion.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SNe absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}.  Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.
One such SN is worth $\gtrsim 25$ galaxies with 0.4~mag absolute magnitude uncertainty.
Secondly,  ZTF today and LSST in the upcoming decade will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands, over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia for which good light curves can be measured. This is a sample size with more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.  Measurement of the velocity field using LSST-discovered
SNe~Ia has been previously quantified by \citet{2011PhRvD..83d3004B,2017JCAP...01..060O}.

The precision in  $f\sigma_8$ derived from the baseline WFD ten-year SN~Ia discoveries has been projected by \citet{2017ApJ...847..128H},
from both their RSD and peculiar velocities.
The results are summarized in Table~\ref{tab:howlett}, where a 5\% distance (0.1~mag) uncertainty is assumed for each supernova.
Their Figure~4b) shows how well these measurements can distinguish between $\gamma=0.42, 0.55, 0.68$
\citep[$f(R)$, GR,  and DGP gravities,][]{2007APh....28..481L} gravities respectively.
SNe~Ia alone can provide a $2\%$ measurement of $f\sigma_8$ at $z<0.3$ redshifts lower than where galaxy and cluster (through the
kinematic S-Z effect)
RSD measurements are sensitive.  A joint galaxy RSD and SN peculiar velocity will be even more powerful.
SNe~Ia peculiar velocities will measure $f\sigma_8$ at $z<0.3$  as well as galaxy surveys will at $z \sim 0.6$ using RSD. 
This can be compared to the current 15\% uncertainty from 6dF \citep{2017MNRAS.471..839A}, 3\% uncertainty projected
for the combined TAIPAN and WALLABY+WNSHS surveys \citep{2017MNRAS.464.2517H} for low-redshift measurements of $f\sigma_8$.
At larger $z>0.3$ redshifts DESI projects 10\% accuracies.

%{\bf Add a SN PV + DESI-like BGS column and/or a PV-only column with Cullan's code? }
% Requires the booktabs if the memoir class is not being used
\begin{table}
   \centering
   %\topcaption{Table captions are better up top} % requires the topcapt package
   \begin{tabular}{@{} lccc @{}} % Column formatting, @{} suppresses leading/trailing space
	\hline
	& RSD & RSD + PV & RSD+PV\\
	Redshift 	 & ($z$-bin) & ($z$-bin)& (cumulative)\\  \hline
      $0.00<z<0.05$   & 66.3 & 13.9 & 13.9\\
     $0.05<z<0.10$            & 24.6     &  7.3 & 6.5\\
     $0.10<z<0.15$      & 14.8  & 5.8 & 4.3\\
     $0.15<z<0.20$      & 10.6  & 5.0 & 3.3\\
      $0.20<z<0.25$     & 8.3  & 4.4 & 2.6\\
     $0.25<z<0.30$  & 6.8  &  4.0 & 2.2\\
      \hline
   \end{tabular}
   \caption{Projected percent uncertainties in $f\sigma_8$ from a 10-year LSST SN survey with 5\% distance uncertainties from
   \citet{2017ApJ...847..128H}. RSD is from clustering, and RSD~+~PV is from joint clustering and peculiar velocities.
   The cumulative column shows the effective uncertainty from combining its and all other shallower redshift bins.}
   \label{tab:howlett}
\end{table}

The projected precisions for LSST-discovered SNe~Ia have a number of interesting features. 
Despite the significant gain in volume and numbers of supernovae, the $f\sigma_8$ uncertainty in increasing redshift bins asymptotes such that 
there is little benefit in going beyond
$z=0.3$.
At $z<0.2$ the constraining power on  $f\sigma_8$ 
comes primarily from peculiar velocities
but by $z=0.3$ RSD is more important.

@@

>>>>

\documentclass{aastex62}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{graphicx}				% Use pdf, png, jpg, or epsÂ§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb,amsmath}
\usepackage{natbib,verbatim}
\usepackage{grffile}
\usepackage{textcase}
\usepackage{hyperref}
\usepackage[overload]{textcase}

@@

\clearpage
\newpage
\setcounter{page}{0}
\pagenumbering{arabic}
\setcounter{page}{1}
\title{Testing Gravity Using Type Ia Supernovae Discovered by Next Generation Wide Field Imaging Surveys}
\author[0000-0001-6315-8743]{A.~G.~Kim}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{G.~Aldering}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{P.~Antilogus}
\affiliation{Laboratoire de Physique Nucl\'eaire et de Hautes Energies, Sorbonne Universit\'e, CNRS-IN2P3, 4 Place Jussieu, 75005 Paris, France}    
\author{S.~BenZvi}
\affiliation{Department of Physics and Astronomy, University of Rochester, Rochester, NY 14627, USA}
\author{S.~Gontcho A Gontcho}
\affiliation{Department of Physics and Astronomy, University of Rochester, Rochester, NY 14627, USA}
\author{R.~Graziani}
\affiliation{Universit\'e Clermont Auvergne, CNRS/IN2P3, Laboratoire de
Physique de Clermont, F-63000 Clermont-Ferrand, France}    
\author{C.~Harper}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{C.~Howlett}
\affiliation{International Centre for Radio Astronomy Research, The University of Western Australia, Crawley, WA 6009, Australia}
\author{D.~Huterer}
\affiliation{Department of Physics, University of Michigan, 450 Church Street, Ann
Arbor, MI 48109, USA }
\author{C.~Ju}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{P.-F.~Leget}
\affiliation{Laboratoire de Physique Nucl\'eaire et de Hautes Energies, Sorbonne Universit\'e, CNRS-IN2P3, 4 Place Jussieu, 75005 Paris, France}    
\author{E.~V.~Linder}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{P.~McDonald}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{J.~Nordin}
\affiliation{Institut fur Physik, Humboldt-Universitat zu Berlin, Newtonstr. 15, 12489 Berlin, Germany}
\author{S.~Perlmutter}
\affiliation{Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\affiliation{Department of Physics, University of California Berkeley, 366 LeConte Hall MC 7300, Berkeley, CA, 94720-7300}
\author{N.~Regnault}
\affiliation{Laboratoire de Physique Nucl\'eaire et de Hautes Energies, Sorbonne Universit\'e, CNRS-IN2P3, 4 Place Jussieu, 75005 Paris, France}    
\author{M.~Rigault}
\affiliation{Universit\'e de Lyon, F-69622, Lyon, France; Universit\'e de Lyon
1, Villeurbanne; CNRS/IN2P3, Institut de Physique Nucléaire de
Lyon, France}    
\author{others}

@@

\begin{abstract}
Cadence wide-field imaging surveys 
in the upcoming decade will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands.  The increase in the number density of SNe~Ia, in parallel with improvements in the standardization of
their absolute magnitudes, now make them competitive probes of the growth of structure.  The peculiar velocity power spectrum
is sensitive to $f\sigma_8$,
the product of the linear growth and amplitude of density perturbations.  Cross-correlation with synergistic galaxy surveys further constrains $f \sigma_8$ and the galaxy bias. 
Thus, in the next decade the peculiar velocities of
SNe~Ia will provide the growth of structure  in the local $z<0.3$ Universe as a powerful test of General Relativity and other models of gravity.
\end{abstract}

\section{}

@@

In the next decade, SNe~Ia will be used as peculiar-velocity probes to measure the the influence of gravity on structure formation within the Universe.
Peculiar velocities are sensitive to the growth of structure through the combination $fD$, where $D$ is  the linear growth factor and
$f \equiv \frac{d\ln{D}}{d\ln{a}}$ is the linear growth rate.  The growth of structure depends on gravity, indeed 
\citet{2007APh....28..481L} find that General Relativity, $f(R)$,  and DGP gravities 
give
$f \approx \Omega_M^\gamma$ with $\gamma=0.55, 0.42, 0.68$ respectively.  Adopting this parameterization for gravity, peculiar velocity
surveys are  sensitive to $fD=\Omega_M^{\gamma} \exp{\left(\int_a^1 \Omega_M^{\gamma} d\ln{a} \right)}$ \citep[see Figure~2 of][to see
$fD$ plotted as a function of redshift for $\Lambda$CDM]{1475-7516-2013-04-031}.

Peculiar velocity surveys have already been  used to measure $fD$, though not at a precision to distinguish gravity models.
 \citet{2017MNRAS.471..839A} use 6dFGS peculiar velocities using the Fundamental Plane method and get distances to
get a 15\% uncertainty in $fD$ at $z\approx 0$ using density and velocity cross-correlations.
The upcoming 
TAIPAN survey will obtain Fundamental Plane galaxies with densities of $n_g \sim 10^{-3}h^3$\,Mpc$^{-3}$,
and the WALLABY+WNSHS surveys will obtain Tully-Fisher galaxies at densities $n_g \sim 2\times 10^{-2} - 10^{-4} h^3$\,Mpc$^{-3}$ from
$z=0-0.1$.
Covering 75\% of the sky to a maximum depth of $z=0.1$, these surveys combined are projected to have 3\% $fD$ uncertainties \citep{2017MNRAS.464.2517H}.
Existing SN~Ia samples
have been used to test and ultimately find spatial correlations in peculiar velocities that may be attributed to the growth of structure
\citep{2008MNRAS.389L..47A,2014MNRAS.444.3926J,2015JCAP...12..033H, 2017JCAP...05..015H}.
For reference, t $z>0.3$ redshifts DESI projects 10\% precision of $fD$ using RSD alone.
%Measurement of the velocity field using LSST-discovered
%SNe~Ia has been quantified by \citet{2011PhRvD..83d3004B,2017JCAP...01..060O}

Two advances in the upcoming decade will make SNe~Ia  important probes of $f\sigma_8$.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields $\sigma_M \gtrsim 0.12$~mag absolute magnitude
dispersion.  However, SNe transmit more information than just the light-curve shape and single color used in current SN models.
Recent studies indicate that with the right data, SNe absolute
magnitudes can be calibrated to $\sigma_M \lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}.  Though not yet
established, it is anticipated that such a reduction in intrinsic dispersion comes with a reduction in the magnitude bias correlated with host-galaxy properties
that is observed using current calibrations.
One such SN is worth $\gtrsim 25$ galaxies with 0.4~mag absolute magnitude uncertainty.
Secondly,  in the upcoming decade cadenced wide-field imaging surveys such as ZTF and LSST
  will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands, over the course of 10-years, LSST will find $\sim150,000$ $z<0.2$,  $\sim520,000$ $z<0.3$ SNe~Ia for which good light curves can be measured. This is a sample size with more galaxies at deeper redshifts than projected by WALLABY and TAIPAN.

A 10-year LSST SN survey can produce interesting $fD$ measurements in redshift bins, as shown in Table~1 of 
\citet{2017ApJ...847..128H}.
Projected percent uncertainties in $f\sigma_8$ from 
\citet{2017ApJ...847..128H}. RSD is from clustering, and RSD~+~PV is from joint clustering and peculiar velocities.
The cumulative column shows the effective uncertainty from combining its and all other shallower redshift bins.

%      $0.00<z<0.05$   & 66.3 & 13.9 & 13.9\\
%     $0.05<z<0.10$            & 24.6     &  7.3 & 6.5\\
%     $0.10<z<0.15$      & 14.8  & 5.8 & 4.3\\
%     $0.15<z<0.20$      & 10.6  & 5.0 & 3.3\\
%      $0.20<z<0.25$     & 8.3  & 4.4 & 2.6\\
%     $0.25<z<0.30$  & 6.8  &  4.0 & 2.2\\

The current sample of SNe~Ia has a low number density compared to Fundamental Plane and Tully-Fisher galaxies.
Nevertheless, their low intrinsic-magnitude uncertainties 
can provide peculiar velocities (expressed equivalently as peculiar magnitudes)
of their host galaxies \citep{2006PhRvD..73l3526H,2011ApJ...741...67D}.

\section{Peculiar Velocity Surveys are Sensitive to the Growth of Structure $\lowercase{f}D$}
In a peculiar-velocity survey of SN~Ia host galaxies, each galaxy has an observed redshift  and a SN-Ia-based distance.
The distance maps onto a cosmological redshift given a model for the background expansion $H_0$-free.
The  (radial) peculiar velocity is obtained from the difference due to the Doppler effect between the observed and cosmological redshifts.
The statistical properties of the peculiar velocities are summarized through their spatial covariance or  their power spectrum $P_{vv}$.

It is standard practice to express peculiar velocities using the formalism created to describe mass overdensities.
In linear theory, the overdensity field can be expressed as
$
\delta (\mathbf{x},a)  = D(a) \delta (\mathbf{x})$,
where $D(a)$ is  the linear growth factor.
The velocity field is related to the density field through the continuity equation as $v \propto fD\delta$,
where $f \equiv \frac{d\ln{D}}{d\ln{a}}$ is the linear growth rate.  The velocity power spectrum $P_{vv} \propto (fD)^2 P_{\delta \delta}$.
Peculiar velocity surveys, combined with a precision measurement of the power spectrum $P_{\delta \delta}$ 
from the CMB, provide a measure of $fD$
(or  equivalently $f\sigma_8$\footnote{
The parameter $\sigma_8$, the  standard deviation of overdensities in 8$h^{-1}$Mpc spheres, is 
commonly used in place of $D$ to normalize the
overall amplitude of  overdensities.}).

The survey's   peculiar velocity signal  can be described by its power spectrum $P_{vv}$.  
Non-GR models can also predict differences in the shape and the redshift-dependence of GR prediction $P_{vv}$,
though this detail is not considered here.
The top panel of Figure~\ref{power:fig} shows the volume-weighted power spectrum $k^3 P_{vv}$
at $z=0$ expected from General Relativity and linear theory. Ranges of $k$ accessed by surveys with different
maximum redshift $z_{max}$ are also indicated, where $k_{min} \approx (2 z_{max})^{-1}$  and 
$k_{max} \approx 0.1$~$h$\,Mpc$^{-1}$ represents  the smallest
scale where  theory predictions are reliable.
Figure~\ref{power:fig} shows  that the bulk of the peculiar velocity signal comes from $0.02 <k< 0.1$~$h$\,Mpc$^{-1}$ modes.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{src/new2.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{Volume-weighted velocity power spectrum $k^3P_{vv}$ at $z=0$ for $\mu=1, 0,5$ as predicted for General Relativity in the linear regime.
The bottom solid horizontal lines show the approximate range of $k$ expected to be used in surveys with corresponding
redshift depths $z_{max}$.
\label{power:fig}}
\end{figure}

The probative power of a specific peculiar-velocity tracer primarily depends on its number density and the precision to which its absolute magnitudes are known.
The current generation of peculiar velocity studies uses $10^3-10^5$ galaxies with Fundamental Plane and Tully-Fisher distances \citep{2008AJ....135.1738M, 2014MNRAS.445.2677S,
2016AJ....152...50T}.  These galaxies have absolute magnitude uncertainties of $\sim 0.4$~mag.

@@

%The results are summarized in Table~\ref{tab:howlett}, where a 5\% distance (0.1~mag) uncertainty is assumed for each supernova.
%Their Figure~4b) shows how well these measurements can distinguish between $\gamma=0.42, 0.55, 0.68$
%\citep[$f(R)$, GR,  and DGP gravities,][]{2007APh....28..481L} gravities respectively.
%SNe~Ia alone can provide a $2\%$ measurement of $f\sigma_8$ at $z<0.3$ redshifts lower than where galaxy and cluster (through the
%kinematic S-Z effect)
%RSD measurements are sensitive.  A joint galaxy RSD and SN peculiar velocity will be even more powerful.
%SNe~Ia peculiar velocities will measure $f\sigma_8$ at $z<0.3$  as well as galaxy surveys will at $z \sim 0.6$ using RSD. 
%
%%{\bf Add a SN PV + DESI-like BGS column and/or a PV-only column with Cullan's code? }
%% Requires the booktabs if the memoir class is not being used
%\begin{table}
%   \centering
%   %\topcaption{Table captions are better up top} % requires the topcapt package
%   \begin{tabular}{@{} lccc @{}} % Column formatting, @{} suppresses leading/trailing space
%	\hline
%	& RSD & RSD + PV & RSD+PV\\
%	Redshift 	 & ($z$-bin) & ($z$-bin)& (cumulative)\\  \hline
%      $0.00<z<0.05$   & 66.3 & 13.9 & 13.9\\
%     $0.05<z<0.10$            & 24.6     &  7.3 & 6.5\\
%     $0.10<z<0.15$      & 14.8  & 5.8 & 4.3\\
%     $0.15<z<0.20$      & 10.6  & 5.0 & 3.3\\
%      $0.20<z<0.25$     & 8.3  & 4.4 & 2.6\\
%     $0.25<z<0.30$  & 6.8  &  4.0 & 2.2\\
%      \hline
%   \end{tabular}
%   \caption{Projected percent uncertainties in $f\sigma_8$ from a 10-year LSST SN survey with 5\% distance uncertainties from
%   \citet{2017ApJ...847..128H}. RSD is from clustering, and RSD~+~PV is from joint clustering and peculiar velocities.
%   The cumulative column shows the effective uncertainty from combining its and all other shallower redshift bins.}
%   \label{tab:howlett}
%\end{table}

Baryonic structures are sensitive to $fD$ through redshift space distortions (RSD), 
which on large scales gives the combination $(b + f \mu^2)D$ where $b$ is the bias between the tracer and dark matter
and $\mu$ gives the angular separation of Fourier modes to the line-of-sight. 
The precision in  $fD$ derived from the baseline LSST WFD ten-year SN~Ia discoveries has been projected by \citet{2017ApJ...847..128H},
from RSD of supernovae and peculiar velocities.

%Baryonic structures and peculiar velocities within the same volume are induced by the same overdensities  making their cross-correlation 
%insensitive to sample variance \citep{2007PhRvL..99h1301G}: galaxy and peculiar velocity surveys provide synergistic constraints more powerful
%than their naive sum.

\section{Probing Gravity}
Our scientific interest is in probing gravity.
The combination $fD$ is sensitive to gravity. The growth rate $f$ describes how fast overdensities increase, which
depends on how fast gravity makes mass move.
General Relativity predicts
$f \approx \Omega_M^\gamma$ with $\gamma=0.55$, whereas other gravity models can be similarly described
with different values of
$\gamma=0.42, 0.68$ for $f(R)$,  and DGP gravities respectively 
\citep{2007APh....28..481L}.

As a probe of gravity, peculiar velocity surveys are sensitive to $fD=\Omega_M^{\gamma} a \exp{\left(\left(\int_0^a \Omega_M^{\gamma} -1\right)  d\ln(a)\right)}$.
Note that $\frac{d(\ln{fD})}{d\gamma} = \ln\Omega + \int \Omega^{\gamma} \ln{\Omega} d\ln{a} \approx -1.68, -0.75,-0.37$ at $z=0,0.5,1.0$ in $\Lambda$CDM,
meaning that surveys at low redshift better measure $\gamma$ given a fixed precent uncertainty in $fD$.
Non-GR models may  also predict a change in the scale-dependence of the growth, such observations provide additional leverage in probing gravity
but are not considered here.

The uncertainties in $\gamma$ for  RSD and peculiar velocity surveys are estimated using a
Fisher matrix analysis
similar to but with lower fidelity to that of \citet{2017ApJ...847..128H, 2017MNRAS.464.2517H}.
The analysis is performed in Fourier space, 
where the Fisher information matrix of parameters $\lambda \in \{\gamma,b\}$ from random Gaussian field with mean zero is given by
\begin{align}
F_{ij} & = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right]\\
& = \frac{\Omega}{8\pi^2} \int_0^{r_{max}}  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\label{fisher:eqn}
\end{align}
 and covariance
\begin{equation}
C(k,\mu)  =
  \begin{bmatrix}
   P_{gg}(k,\mu) + \frac{1}{n_g} &
   P_{vg}(k,\mu)  \\
   P_{vg}(k,\mu)  &
  P_{vv}(k,\mu) + \frac{\sigma^2}{n}
   \end{bmatrix}.
\label{cov:eq}
\end{equation}
The uncertainty in $\gamma$ is $\sqrt{\left(F^{-1}\right)_{\gamma,\gamma}}$.

The survey parameters that enter the calculation are the solid angle $\Omega$, depth of the survey $r_{max}$ (or $z_{max}$), the intrinsic magnitude dispersion 
through $\sigma \approx (\frac{5}{\ln{10}} \frac{1+z}{z})^{-1} \sigma_M$, and the survey duration $t$ through $n=\epsilon rt$ where $r$ is the observer SN~Ia rate
and $\epsilon$ is the sample-selection efficiency.
An analysis in which the RSD and velocity surveys are treated independently (e.g.\ they do not cover the same volume) is considered by
setting the off-diagonal elements of $C$ to zero.

We consider fiducial peculiar velocity surveys for a range of redshift depths $z_{max}$, with survey durations or $t=2$ and 10~years, and with or without
cross-correlations accounted for.
The other survey parameters
$\Omega=3$, $\epsilon=0.65$, $\sigma_M=0.08$~mag are fixed.
A TAIPAN-like
galaxy survey with $b=1.2$ and number density $n_g=10^{-3}h^3$\,Mpc$^{-3}$ is used for the RSD measurement.
The uncertainties in $\gamma$ are shown in Figure~\ref{var:fig}, which shows a number of interesting features:
\begin{itemize}
\item Even shallow, two-year surveys give gammas for 
\end{itemize}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{src/var.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{
\label{var:fig}}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.47\textwidth]{src/dvardz.png}
\includegraphics[width=0.47\textwidth]{src/dvardlnt.png}
%\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{
\label{dvar:fig}}
\end{figure}

%The growth of structure depends on the expansion history of the Universe, the nature and density of its contents, and gravity. 
%It is therefore a powerful probe of cosmology and dark energy.  Growth of structure can be measured from the light-emittting structures
%in the Universe and from the peculiar velocities of test masses therein.
%Peculiar velocities are the motions, on top of the cosmological expansion, caused by the gravitational attraction
%and repulsion of density inhomogeneities in the Universe.  The peculiar velocity of an object with a known absolute magnitude
%can be determined from its observer-frame magnitude and redshift. For a given background cosmology, the observed magnitude provides
%an estimate of the cosmological redshift, the peculiar velocity is then the difference between the cosmological and CMB-frame redshifts.

%Correlations between peculiar velocities are sensitive to  $f\sigma_8 (H_0 d_L)^{-1}$.
%Although the velocities may be measured off of biased tracers, the tracers' dynamics are driven by all mass (including dark matter) and so are bias-free.
%As
%mentioned earlier, peculiar velocities are measured relative to the background cosmological expansion that
%leads to the dependence on $H_0 d_L$. (Note that the combination $H_0 d_L$ is independent of $H_0$.)

%
%The projected precisions for LSST-discovered SNe~Ia have a number of interesting features. 
%Despite the significant gain in volume and numbers of supernovae, the $f\sigma_8$ uncertainty in increasing redshift bins asymptotes such that 
%there is little benefit in going beyond
%$z=0.3$.
%At $z<0.2$ the constraining power on  $f\sigma_8$ 
%comes primarily from peculiar velocities
%but by $z=0.3$ RSD is more important.

@@

\newpage
We are interested in
$$F^{-1}_{00,\alpha} = \frac{F_{11,\alpha}}{F_{00}F_{11}-F_{01}^2} - \frac{F_{11}}{(F_{00}F_{11}-F_{01}^2)^2}\left(F_{00,\alpha}F_{11}+
F_{00}F_{11,\alpha} - 2 F_{01}F_{01,\alpha}\right).$$

\begin{align}
F_{ij} &  = \frac{\Omega}{8\pi^2} \int_0^{r_{max}}  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk\,dr
\end{align}

$$F_{ij,z} =  \frac{\Omega}{8\pi^2} \frac{dr}{dz} r^2  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1}  k^2 \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right] d\mu\,dk$$

$$F_{ij,s} =  \frac{\Omega}{8\pi^2} \int_0^{r_{max}}  \int_{k_{min}}^{k_{max}}  \int_{-1}^{1} r^2 k^2 \text{Tr}\left[
C^{-1}_{,s} \frac{\partial C}{\partial \lambda_i} C^{-1}\frac{\partial C}{\partial \lambda_j} +
C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}_{,s}\frac{\partial C}{\partial \lambda_j} 
\right] d\mu\,dk\,dr$$
